Supplement History of Cheminformatics as part of Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-24 rebuttal.tex

\comment{The manuscript is poor in relevant citations with none of the
  classical references provided, such as Morgan 1965, Weininger 1988,
  Johnson and Maggiora 1990, Barnard 1993, Willett et al. 1998, etc.}
\response{The text has been updated to include the ``classic''
  references. In addition the supplemental history document lists a
  number of milestones in cheminformatics and their relevant references.}

\comment{Molecular docking is also not mentioned, but is regarded as
  one of the key methods applied in cheminformatics. Docking should be
  included, along with other molecular modelling methods such as
  pharmacophores, shape searching, molecular dynamics, as a distinct
  section.}
\response{These methods are mentioned on page 5 (column
  1). However, we have chosen not to place a more in depth discussion
  of these methods in their own section in the main document for a few
  reasons. First, due to space limitations this would require us to
  compress or drop other sections which we believe to be more
  relevant. Second, we are of the view that docking, for example, is
  not really a cheminformatics technique, though it makes extensive
  use of cheminformatic methods. Instead we place it under molecular
  modeling. We realize that this is a somewhat subjective
  ``classification'' of methods. However we do make explicit note of
  these methods in the supplemental document where we make a
  distinction between cheminformatics and related subject areas such
  as quantum mechanics and molecular modeling.}

\comment{Proteochemometric methods are mentioned a couple of times but
  these are largely un-noticed by the community currently. I do not
  think that it is necessarily appropriate to include these as
  tried-and-tested methods in an over-arching review such as this.}
\response{It is true that proteochemometric methodologies
  might not be as well known as say docking or pharmacophore modeling.
  Though, what is considered as proteochemometric, is in docking known as 
  'interaction fingerprint', e.g. for Kinase databases, and in the field of HIV the
  term 'phenotypic modeling' has a long tradition.
  We choose, along with multiple reviews and the books written by 
  Qing Yan (Pharmacogenomics) and Kubinyi/M\"{u}ller (Chemogenomics), 
  that the term proteochemometrics is worth being mentioned, since it unifies the 
  (statistical) mining/QSAR techniques that do explicitly include multiple ligand 
  activities and multiple receptor features.\\
  Finally, we also agree with reviewer number one that proteochemometric modeling
  plays an important role in translational medicine, e.g. HIV-HAART therapy. 
  A paragraph and references about translational medicine was added to the 
  historic supplement.}

### 2011-11-28 OpenQuestions.tex
included JVDs suggestions

This question encompasses the open questions of \emph{approximately counting the number of stereoisomers, or the number of tautomers, of a chemical formula}.  Goldberg and Jerrum designed an algorithm which, given a chemical formula, would output an isomer of that formula chosen uniformly at random~\cite{RandomlySampling}.  Perhaps the main theoretical obstacle to overcome is that molecules tend to be \emph{chiral} (that is, they have a ``handedness'' or 3D orientation), whereas traditional graph theory (and the Goldberg/Jerrum algorithm) treats graphs with identical vertices and edges as isomorphic.  So part of the question could be rephrased as, ``Given a labeled vertex set, count the number of structures that are identical on that vertex set, except that they differ in their 3D orientation.''  Mathematical chemists have designed measures of molecular chirality~\cite{ChiralityMeasures}. A resolution of this problem may require connecting graph enumeration algorithms to knot theory or other topics in topology~\cite{TopologicalLook}.
%
\item \emph{Write code that approximately counts the number of conformers of a chemical formula.}

While this is related to the previous question, it is conceivable that one could answer this question without solving the theoretical problem.  For example, as we note in the main article, the commercial software MOLGEN can approximately enumerate structures that satisfy given spectral data, even though related theoretical enumeration problems appear wide open.
%
\item \emph{Design and implement efficient subgraph isomorphism algorithms for useful special cases of molecular graphs}.

@@

\section*{Databases \& software engineering}
\begin{enumerate}
\item \emph{Large scale conformational databases}.

>>>>

  This question encompasses the open questions of \emph{approximately
    counting the number of stereoisomers, or the number of tautomers,
    of a chemical formula}.  Goldberg and Jerrum designed an algorithm
  which, given a chemical formula, would output an isomer of that
  formula chosen uniformly at random~\cite{RandomlySampling}.  Perhaps
  the main theoretical obstacle to overcome is that molecules tend to
  be \emph{chiral} (that is, they have a ``handedness'' or 3D
  orientation), whereas traditional graph theory (and the
  Goldberg/Jerrum algorithm) treats graphs with identical vertices and
  edges as isomorphic.  So part of the question could be rephrased as,
  ``Given a labeled vertex set, count the number of structures that
  are identical on that vertex set, except that they differ in their
  3D orientation.''  Mathematical chemists have designed measures of
  molecular chirality~\cite{ChiralityMeasures}. A resolution of this
  problem may require connecting graph enumeration algorithms to knot
  theory or other topics in topology~\cite{TopologicalLook}.
%
\item \emph{Design and implement efficient subgraph isomorphism
    algorithms for useful special cases of molecular graphs}.

@@

\section*{Databases \& software engineering}
\begin{enumerate}

\item \emph{Real time substructure searching in massive chemical
    databases}.

  Chemical databases have grown tremendously in size. A common task in
  such databases is substructure searching. However most cases of very
  large structure databases do not support substructure in real time
  (something that would allow applications such as
  ``type-ahead''). How can one enable such rapid substructure searches
  on massive ($> 10^9$ molecules) collections of structures? This
  problem has various aspects:
  \begin{itemize}
  \item What type of indexing schemes will support rapid substructure searches?
  \item How can other molecular properties be included in substructure
    searchs that allow ranking of results, all the while maintaining
    real-time response (i.e. $< 1$ sec)?
  \item What type of database architectures are required for this
    scale of structure searching?
  \end{itemize}

\item \emph{Large scale conformational databases}.

### 2011-11-28 supplement_history.tex
Update rebuttal. Added ref and updated history
Added some more comments and fixed typos.
Adding book references for Chemo-/Pharmacogenomics and relation to translational medicine.
included JVDs suggestions

to highlight some historic milestones.  In addition, we also discuss

@@

@@

  efforts \cite{Dudley:2011fk,Swamidass:2011uq}.

@@

From a more technical viewpoint, the first chemical graphs were drawn
by the Scottish chemist Willam Cullen in 1758
\cite{bonchevrouvrey1991}, who initially called them affinity graphs.
After developing the concept of bonds between atoms (Couper, 1858) the
first chemical graphs occured in publications of Brown in 1864 and
Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.

>>>>

highlight some historic milestones.  In addition, we also discuss

@@

\item ``\textit{The set of approaches to computer-aided drug design
    that do not rely on the 3D structure of the [protein] target}''
  John Van Drie, 2011 (personal communication)

@@

  efforts \cite{Dudley:2011fk,Swamidass:2011uq}. Related topics and
  books are also Pharmacogenomics \cite{yan2008pharmacogenomics} and
  Chemogenomics \cite{kubinyi2004chemogenomics}.

@@

From a more technical viewpoint, the first chemical graphs were drawn
by the Scottish chemist Willam Cullen in 1758
\cite{bonchevrouvrey1991}, who initially called them affinity graphs.
After developing the concept of bonds between atoms (Couper, 1858) the
first chemical graphs occured in publications of Brown in 1864 and
Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}. The notion of
mathematical chemistry in general was discussed by Helm in 1897
\cite{Helm:1897ys} and the reader is referred to Balaban
\cite{Balaban:2005zr} for a historical overview of this field.
